Executive Constraints and Civil Conflict Onset

### 2020-07-03 paper.Rmd
Draft edits

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of conflict. In that time period, upwards to
two-thirds of all countries have experienced at least one minor
insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging from
motivation-based arguments --- such as, for example, the role of
horizontal inequalities [@Cederman2011] --- to opportunity-based
explanations centered on economic factors [@Collier2004; @Collier2009]
as well as the political and structural conditions underlying state
weakness [@Hegre2001; @Fearon2003]. This latter strain of research in
particular has spurred extensive work into examining the
distinguishing features of weak states and whether political
institutions play a significant role in the risk of civil conflict.
Several studies in this vein have argued in favor of institutions by
utilizing disaggregated measures to also investigate the specific
types of institutions relevant to the study of civil conflict
[@Taydas2010; @Hegre2015a; @Bartusevic2018].

This paper extends that work by taking a disaggregated approach to
studying the effect of executive constraints on the risk of civil
conflict onset. The motivation to specifically focus on executive
constraints is two-fold: first, the executive is often times one of
the key decision-makers in a state's regime. Institutions that provide
a check against the types of actions which may exacerbate the risk of
rebellion should be relevant to the study of civil conflict. In doing
we can further our understanding of how political institutions shape
the incentives for violence, especially through the influence of
institutions on state behaviour.  Second, the link between executive
constraints and civil conflict is relatively understudied. The
previous literature that exists is fairly limited and contradictory in
terms of results [@Getmansky2013; @Fjelde2009].

The principle finding is that there is tentative evidence that
executive constraints do matter and that there is a negative
relationship between the latent level of constraints on the executive
and the risk of civil conflict onset. However, this relationship
largely disappears for low-income country-years when an interaction
term with GDP per capita is included. If the primary theoretical
mechanism involves the impact of executive constraints on rebel
mobilization --- an unchecked executive may be more likely to engage
in indiscriminate violence which could lead to a loss of legitimacy as
well as the ability of rebel groups to offer local security guarantees
to attract recruits --- then the incentives for violence in low-income
country-years may be so great that it does not matter how constrained
the executive may or may not be.

We should not be overconfident with these results, however. Similar to
most other studies examining conflict onset, the model is a poor fit
to the underlying data. Modeling civil conflict is a difficult task,
especially when trying to capture the dynamics of conflict with a
standard time series cross-country regression. The results will
require verification by additional studies using alternative
methodological approaches.

In what follows, a deeper justification for why political institutions
are relevant to the study of civil conflict together with an overview
of related empirical literature will be presented. The theoretical
mechanisms linking executive constraints to conflict onset will then
be discussed followed by an empirical analysis for which a latent
estimate for executive constraints will be estimated and used in a
cross-country time series logistic regression predicting conflict
onset.

@@

Unsurprisingly, insofar as the role of institutions has been examined
by previous literature the focus has largely been on the broader
notion of democracy, which at a minimum is a system regulating access
to political power through popular representation determined by free
and competitive elections [@Schumpeter1962]. Democracy serves as a
proxy for a number of normatively desirable institutions and has been
hypothesised to have a pacifying domestic effect according to two
predominant mechanisms. The first is centered on the concept of
relative deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence [@Gurr1970]. [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means --- for example, through the regular occurrence of
free and fair elections [@Hegre2014]. The second mechanism assumes
that violent conflict is costly and that there exists a bargained
solution preferable to fighting, assuming that actors are rational and
able to internalize the cost of violence [@Fearon1995]. One of the
reasons why a bargained solution may be impossible is the risk of
future shifts in the distribution of power. More specifically, a
future change in favor of the relative strength of one of the parties
can provide an incentive to later renege. This influences the present
degree to which concessions can be offered and parties are willing to
credibly commit [@Powell2006]. Within the domestic sphere, democracy
offers a possible solution by formalising the regular contest of power
with full enfranchisement, thereby providing a *de jure* guarantee for
the balance of power [@Origins2006], and should, likewise, decrease
the likelihood of civil conflict.

@@

In contrast, studies emphasizing the importance of political and
economic opportunities for rebellion in understanding when and where
conflicts break out, such as @Fearon2003 and @Collier2009, find
democracy to be insignificant. According to their theoretical
expositions, democracy should be associated with higher levels of
political rights, respect for ethnic and religious minorities, and
less repression, thus being a proxy for the latent level of political
discontent against the state. However, the argument follows that even
if the number of political grievances vary across regime types, they
are still ubiquitous enough to be a poor explanation for why conflict
breaks out in only a relatively small subset of countries. Instead,
the determining factors for conflict onset are the structural
conditions conducive to rebellion.[^opportunities]

[^opportunities]: For Fearon and Laitin these structural conditions
    are related to the features which make rebellion militarily
    viable.  Specifically, there's a higher risk of conflict onset
    where state capacity is weak, principally understood in terms of
    state revenue and its impact on policing and counterinsurgency
    operations; large impoverished populations; and insurgents having
    a terrain advantage. Collier *et al* start from the premise that
    war is expensive and the primary impediment to a rebellion is its
    financial feasibility relative to the strength of the state. On
    the basis of their results they go so far as provide support to
    the proposition, "where a rebellion is financially and militarily
    feasible it will occur." Both approaches, however, reach the same
    conclusion concerning the empirical evidence surrounding the role
    of democracy.

To provide a more nuanced picture, an ongoing strain of research has
been dedicated to disaggregating measures of democracy into either
finer regime type categories or specific measures of individual
political institutions. The theories offered tend to follow a general
formula whereby formal lack or poor quality of some institution(s),
measured either directly or proxied by regime type, undermines the
capacity of the state to prevent armed insurgency and/or incentivises
opposition violence by restricting the opportunities for political
participation and increasing political discontent. For example,
@Carey2007 examines the executive selection process of Sub-Saharan
Africa and argues that executive elections, even when they are of poor
quality, reduce the risk of large-scale violent dissent. @Fjelde2010
breaks down authoritarian regime types in Africa attributing the
significant variation in conflict risk to the differing institutional
configurations affecting a regime's ability to coerce or co-opt the
opposition. The conclusion is that previous indeterminate results
regarding the link between political institutions and conflict risk
may have been hamstrung by aggregate measures which poorly reflect the
within-group variation of political regimes. In a larger sample
including democratic polities, @Goldstone2010 create a regime
classification scheme based on measures for openness of executive
recruitment and competitiveness of political participation, as well as
factionalism in order to distinguish between partial democracies. They
offer a theory of civil conflict occurring in countries where elite
divisions undermine the capacity of the state to defeat an insurgency
and claim that a relatively parsimonious model including their regime
classification offers better predictive accuracy than previous
analyses and provides evidence that political institutions are a
dominant driving factor for both political instability and civil
conflict. Finally, @Bartusevic2018 construct a regime typology focused
specifically on electoral features, which are argued form the
essential core of the concept of democracy. Where there are
constraints on electoral contestation, the authors contend that such
constraints restrict the opportunities for non-violent political
participation thereby legitimising anti-government violence and
self-selecting recruits into rebel groups willing to engage in armed
conflict. Evidence is presented which the authors claim supports not
just the significance of regime type, but a negative monotonic
relationship between democracy and the risk of civil conflict.

Studies utilising individual measures beyond regime classification
have been largely focused on the concept of governance. For instance,
@Taydas2010 examine institutional quality as a means to capturing a
regime's governing capacity which is theorised to be linked to
conflict onset both in terms of political discontent --- a regime's
inability to govern effectively and respond to public needs leading to
grievances and a loss in legitimacy --- and political opportunities
for rebellion --- poor institutional quality undermines the state's
ability to handle a nascent, violent uprising. Specifically, they
assess corruption, rule of law, and bureaucratic quality and present
evidence that all three are negatively related to civil conflict
onset. Similarly, @Hegre2015a argue that the presence of formal
representative institutions does not fully capture the extent to which
a regime is capable of realizing policy goals benefiting the
public. Poor quality of government can therefore undermine the
legitimacy of a regime when citizens identify the state as responsible
for relative deprivation. In their empirical analysis, the authors
separate informal institutions related to good governance from formal
democratic institutions and find both important for reducing the
incidence of civil conflict, or in this specific case the recurrence
of conflict.

@@

[^ODonnell]: Adapted from the definition of horizontal accountability
    offered by @ODonnell1998 who further argues that horizontal
    accountability can potentially be a distinguishing feature between
    consolidated democracies and newly democratic states, the latter
    of which may have free and open elections, but are considerably
    weaker in upholding liberal notions of individual freedom against
    state power and republican notions of representation. In this way,
    executive constraints can possibly aid in discriminating the types
    of partially democratized regimes which may be at a heightened
    risk for civil conflict.

@@

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset. This proposition can be further extended if state weakness is a
primary driver of civil conflict and the relative capacity of the
state is effected by the levels of executive constraints, as depicted
in the first theoretical pathway regarding executive looting, then
state capacity could have a mediating affect on the relationship
between executive constraints and conflict onset. Another extension
would be if executive constraints affect the mobilization potential
for rebel groups, then there could also be a confounding effect from
the relative threshold for mobilization. Finally, the relationship
could still be confounded by the overall opportunity for rebellion ---
the degree of executive constraints may not be particularly relevant
in relatively stable countries lacking the conditions conducive to
carrying out a rebellion. Regardless of conditionalities, contrary
results would indicate either that executive constraints are
indeterminate to the risk of civil conflict or possibly that there is
even a positive relationship since constrained executives may be less
able to effectively deal with violent challenges due to risk aversion
to violence and/or from the slowdown of having to seek approval from a
plurality of stakeholders before committing to decisive action.

@@

I select a weakly  informative prior for $\eta$ that is skeptical of
large values given that $\nu$ is on the log-odds scale. For $\rho$ I
use an informative prior that constrains an overwhelming majority of
the prior probability mass to a *length-scale* between 4 and 10, which
provides a strong smoothing effect and aids in computation by
avoiding unreasonable values [@Betancourt2017].
\begin{align*}
    \eta & \sim \mathsf{HalfNormal}(0, 1) \\
    \rho & \sim \mathsf{InvGamma}(15.5031, 78.2992)
\end{align*}

@@

[^thresholds]: Thresholds relate to how the continuous latent trait
    for a given variable is mapped on to the ordinal answer categories
    presented on the survey. Due to differential item functioning,
    experts may perceive the magnitude of difference between answer
    items differently. For example, for the variable *v2elvotbuy*,
    election vote buying, experts may vary in how they interpret the
    difference between a 2 ("Restricted Vote Buying") and a 3 ("Almost
    No Vote Buying"). These two concepts, rater reliability and
    varying rater thresholds, correspond respectively to the
    *discrimination* and *difficulty* parameters in classical IRT
    models.

@@

In the case of the legislative variables there is the added
issue of missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves unbounded latent variables, minimal values cannot be
imputed without a nonlinear transformation to a bounded scale. I thus
take a cue from *v2x_horacc* and include into the model a dichotomized
version of *v2lgbicam*, how many chambers does the legislature
contain, denoted by the Bernoulli distributed variable, $w$, where
country-years with a sitting legislature are given a $1$ and
country-years where the legislature is missing or has been suspended,
$0$. The probability of a legislature being present, $\pi_i$, is then
a linear function of the latent factor, $\theta$, transformed by the
inverse logit such that $\pi_i \in [0, 1]$. The practical effect is
that where there is data for the legislative manifest variables, the
latent factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables. The resulting model specification, including
priors, is as follows:
\begin{align*}
    w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
    \pi_i & = \mathsf{logit}^{-1}(\eta + \delta \theta_i) \\
    \eta & \sim \mathsf{Normal}(0, 5) \\
    \delta & \sim \mathsf{LogNormal}(0, 1)
\end{align*}

@@

As mentioned previously, to control for time dependence --- past and
current civil conflicts can impact the present risk of conflict onset
as well as the current level of executive constraints -- I include the
number of peace years since the last conflict incident.[^time] I also
include a dummy variably for ongoing high intensity conflict since
devastating civil ways may have a particularly destabilizing
effect.[^intensity] To account for spillover effects, I include a
dummy variable for whether a neighbouring country is currently
experiencing a civil conflict.[^geo] Finally, one notable lack is a
general measure of political instability, which I omit since it could
be highly correlated with executive constraints and would potentially
bias results, at least for years of crisis where the risk of civil
conflict is highest. I do, however, include a dummy variable for the
independence year since newly sovereign states may be especially
vulnerable.

@@

Table \ref{tab:fa} and figure \ref{fig:fpars} summarise the posterior
draws for the measurement model parameters with 80 percent quantile
intervals. According to the results, the factor loadings, $\lambda$,
for each manifest variable are fairly comparable, with the exception
of the lower value for *v2lgfunds* ("legislature controls its own
resources") and to a lesser extent *v2lgqstexp* ("legislature
questions executive branch officials in practice"). Relative to the
other manifest variables, both items are less informative in inferring
the values of the latent factor --- a $\lambda$ of zero would indicate
that the latent factor does not explain any of the variation in the
manifest variable. As for the variable specific intercepts, $\gamma$,
they are largely clustered around zero as to be expected given that
the manifest variables were standardized prior to the model run. Where
deviations do occur, most notably with several of the legislative
variables, it is because of the measurement error component of the
model shifting the mean for a specific variable away from zero. For
example, after adjusting for measurement error, the mean for
$v2lgfunds$ is on average slightly higher than the remaining manifest
variables, hence the negative intercept.

(ref:fa)

Executive constraints measurement model parameter
    estimates with 80 percent posterior credible intervals

@@

```{r fig.width = 7, fig.height = 5, fig.cap = "\\label{fig:theta} (ref:theta)"}
theta_summarised <- post_summarise(theta, probs = c(0.10, 0.5, 0.90)) %>%
    rename(codelow = `10%`, median = `50%`, codehigh = `90%`)

@@

(ref:cap) Executive constraints in 2000 versus 2010 for several
    selected countries.

```{r fig.cap = "\\label{fig:tie} (ref:cap)"}
targets <- c("Algeria", "Argentina", "Armenia", "Chile", "Colombia",
             "Egypt", "Georgia", "Ghana", "Hungary", "Indonesia",
             "Iran", "Israel", "Japan", "Kenya", "Mexico", "Poland",
             "Portugal", "Romania", "Rwanda", "Turkey", "Tunisia",
             "United States of America", "Zimbabwe")

@@

levels <- filter(df, year == 2010) %>%
    arrange(median) %$%
    country_name

df$country_name <- factor(df$country_name, levels = levels)

ggplot(df, aes(country_name, median, colour = year)) +
    geom_errorbar(aes(ymax = codehigh, ymin = codelow),
                  position = position_dodge(width = 0.5)) +
    theme_tufte() +
    scale_color_colorblind() +
    theme(axis.title.y = element_blank(),
          legend.title = element_blank()) +
    ylab("Theta") +
    coord_flip()

@@

(ref:reg) 80 percent posterior credible intervals for regression
    slopes

```{r reg, fig.height = 3}
reg <- read_post("posteriors/full_model/reg_posteriors.csv")

>>>>

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of state-based conflict. In that time period,
upwards to two-thirds of all countries have experienced at least one
minor insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging from
motivation-based arguments --- such as, for example, the role of
horizontal inequalities and ethnic identities [@Cederman2011] --- to
opportunity-based explanations centered on economic factors
[@Collier2004; @Collier2009] as well as the political and structural
conditions underlying state weakness [@Hegre2001; @Fearon2003]. This
latter strain of research in particular has spurred extensive work
into examining the distinguishing features of weak states and whether
political institutions play a significant role in the risk of civil
conflict. Several studies in this vein have argued in favor of
institutions by investigating, for example, aspects of governance,
institutional quality, and opportunities for political participation
[@Fjelde2009; @Taydas2010; @Hegre2015a; @Bartusevic2018].

This paper extends that work by taking a disaggregated approach to
studying political institutions by focusing specifically on the effect
of executive constraints on the risk of civil conflict onset. The
motivation to focus on executive constraints is two-fold: first, the
executive is often times one of the key decision-makers in a state's
regime. Institutions that provide a check against the types of actions
which may exacerbate the risk of rebellion should be relevant to the
study of civil conflict. Such an approach can further our
understanding of how specific political institutions may shape the
incentives for violence, especially through the influence on state
behaviour. Second, the link between executive constraints is
relatively understudied. The previous literature that exists is fairly
limited and contradictory in terms of results.

The principle finding of this paper is that there is tentative
evidence that executive constraints do matter and that there is a
negative relationship between the latent level of constraints on the
executive and the risk of civil conflict onset. However, this
relationship largely disappears for low-income country-years when an
interaction term with GDP per capita is included. If the primary
theoretical mechanism involves the impact of executive constraints on
rebel mobilization --- an unchecked executive may be more likely to
engage in indiscriminate violence which could lead to a loss of
legitimacy as well as the ability of rebel groups to offer local
security guarantees to attract recruits --- then the incentives for
violence in low-income country-years may be so great that it does not
matter how constrained the executive may or may not be.

We should not be overconfident with these results, however. Similar to
most other studies examining conflict onset, the presented models are
a poor fit to the underlying data. Modeling civil conflict is a
difficult task, especially when trying to capture the dynamics of
conflict with a standard time series cross-country regression. The
results will require verification by additional studies using
alternative methodological approaches.

In what follows, a deeper justification for why political institutions
are relevant to the study of civil conflict together with an overview
of related empirical literature will be presented. The theoretical
mechanisms linking executive constraints to conflict onset will then
be discussed followed by an empirical analysis using a bayesian latent
variable model to estimate executive constraints from the basis of
several observed manifest variables. This latent measure will then be
used to predict civil conflict onset using a cross-country time series
logistic regression.

@@

Unsurprisingly, insofar as the role of institutions on the likelihood
of civil conflict has been examined by previous literature the focus
has largely been on the broader notion of democracy, which at a
minimum is a system regulating access to political power through
popular representation determined by free and competitive elections
[@Schumpeter1962]. Democracy serves as a proxy for a number of
normatively desirable institutions and has been hypothesised to have a
pacifying domestic effect according to two predominant mechanisms. The
first is centered on the concept of relative deprivation, or
inequalities in value expectations and value capabilities, as a
primary source of discontent that can motivate collective violence
[@Gurr1970]. [^gurr] Democracy is thought to decrease the amount of
political discontent by providing avenues for free expression and
adjudicating group conflicts through non-violent institutional means
--- for example, through the regular occurrence of free and fair
elections [@Hegre2014]. The second mechanism assumes that violent
conflict is costly and that there exists a bargained solution
preferable to fighting, assuming that actors are rational and able to
internalize the cost of violence [@Fearon1995]. One of the reasons why
a bargained solution may be impossible is the risk of future shifts in
the distribution of power. More specifically, a future change in favor
of the relative strength of one of the parties can provide an
incentive to later renege. This influences the present degree to which
concessions can be offered and parties are willing to credibly commit
[@Powell2006]. Within the domestic sphere, democracy offers a possible
solution by formalising the regular contest of power with full
enfranchisement, thereby providing a *de jure* guarantee for the
balance of power [@Origins2006], and should, likewise, decrease the
likelihood of civil conflict.

@@

In contrast, studies emphasizing the importance of political and
economic opportunities for rebellion in understanding when and where
conflicts break out, such as @Fearon2003 and @Collier2009, find
democracy to be insignificant. According to their theoretical
expositions, democracy should be associated with higher levels of
political rights, respect for ethnic and religious minorities, and
less repression, thus being a proxy for the latent level of political
discontent against the state. However, the argument follows that even
if the number of political grievances vary across regime types, they
are still ubiquitous enough to be a poor explanation for why conflict
breaks out in only a relatively small subset of countries at any given
point in time. Instead, the determining factors for conflict onset are
the structural conditions conducive to rebellion.[^opportunities]

[^opportunities]: For Fearon and Laitin these structural conditions
    are related to the features which make rebellion militarily
    viable.  Specifically, there's a higher risk of conflict onset
    where state capacity is weak, principally understood in terms of
    state revenue and its impact on policing and counterinsurgency
    operations; large impoverished populations; and insurgents having
    a terrain advantage. Collier *et al* start from the premise that
    war is expensive and the primary impediment to a rebellion is its
    financial feasibility relative to the strength of the state. On
    the basis of their results they go so far as provide support to
    the proposition, "where a rebellion is financially and militarily
    feasible it will occur" [@Collier2009]. Both approaches, however,
    reach the same conclusion concerning the empirical evidence
    surrounding the role of democracy.

To provide a more nuanced picture, an ongoing strain of research has
been dedicated to disaggregating measures of democracy into either
finer regime type categories or specific measures of individual
political institutions. The theories offered tend to follow a general
formula whereby formal lack or poor quality of some institution(s),
measured either directly or proxied by regime type, undermines the
capacity of the state to prevent armed insurgency and/or incentivises
opposition violence by restricting the opportunities for political
participation and increasing political discontent. For example,
@Carey2007 examines the executive selection process of Sub-Saharan
Africa and argues that executive elections, even when they are of poor
quality, reduce the risk of large-scale violent dissent. @Fjelde2010
breaks down authoritarian regime types in Africa attributing the
significant variation in conflict risk to the differing institutional
configurations affecting a regime's ability to coerce or co-opt the
opposition. The conclusion is that previous indeterminate results
regarding the link between political institutions and conflict risk
may have been hamstrung by aggregate measures which poorly reflect the
within-group variation of political regimes.

In a larger sample including democratic polities, @Goldstone2010
create a regime classification scheme based on measures for openness
of executive recruitment and competitiveness of political
participation, as well as factionalism in order to distinguish between
partial democracies. They offer a theory of civil conflict occurring
in countries where elite divisions undermine the capacity of the state
to defeat an insurgency and claim that a relatively parsimonious model
containing their regime classification offers better predictive
accuracy than previous analyses and provides evidence that political
institutions are a dominant driving factor for both political
instability and civil conflict. Finally, @Bartusevic2018 construct a
regime typology focused specifically on electoral features, which are
argued form the essential core of the concept of democracy. Where
there are constraints on electoral contestation, the authors contend
that such constraints restrict the opportunities for non-violent
political participation thereby legitimising anti-government violence
and self-selecting recruits into rebel groups willing to engage in
armed conflict. Evidence is presented which the authors claim supports
not just the significance of regime type, but a negative monotonic
relationship between democracy and the risk of civil conflict.

Studies utilising individual measures beyond regime classification
have been largely focused on the concept of governance. For instance,
@Taydas2010 examine institutional quality as a means to capturing a
regime's governing capacity which is theorised to be linked to
conflict onset both in terms of political discontent --- a regime's
inability to govern effectively and respond to public needs leading to
grievances and a loss in legitimacy --- and political opportunities
for rebellion --- poor institutional quality undermines the state's
ability to handle a nascent, violent uprising. Specifically, they
assess corruption, rule of law, and bureaucratic quality and present
evidence that all three are negatively related to civil conflict
onset. Similarly, @Hegre2015a argue that the presence of formal
representative institutions does not fully capture the extent to which
a regime is capable of realizing policy goals benefiting the
public. Poor quality of government can therefore undermine the
legitimacy of a regime when citizens identify the state as responsible
for relative deprivation. In their empirical analysis, the authors
separate informal institutions related to good governance from formal
democratic institutions and find both important for reducing the
incidence of recurring civil conflict.

@@

[^ODonnell]: This is adapted from the definition of horizontal
    accountability offered by @ODonnell1998 who further argues that
    horizontal accountability can potentially be a distinguishing
    feature between consolidated democracies and newly democratic
    states, the latter of which may have free and open elections, but
    are considerably weaker in upholding liberal notions of individual
    freedom against state power and republican notions of
    representation. In this way, executive constraints can possibly
    aid in discriminating the types of partially democratized regimes
    which may be at a heightened risk for civil conflict.

@@

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset. This proposition can be further extended: if state weakness is a
primary driver of civil conflict and the relative capacity of the
state is effected by the levels of executive constraints, as depicted
in the first theoretical pathway regarding executive looting, then
state capacity could have a mediating affect on the relationship
between executive constraints and conflict onset. Another extension
would be if executive constraints affect the mobilization potential
for rebel groups, then there could also be a confounding effect from
the relative threshold for mobilization. Finally, the relationship
could still be confounded by the overall opportunity for rebellion ---
the degree of executive constraints may not be particularly relevant
in relatively stable countries lacking the conditions conducive to
carrying out a rebellion. Regardless of conditionalities, contrary
results would indicate either that executive constraints are
indeterminate to the risk of civil conflict or possibly that there is
even a positive relationship since constrained executives may be less
able to effectively deal with violent challenges due to risk aversion
to violence and/or from the slowdown of having to seek approval from a
plurality of stakeholders before committing to decisive action.

@@

I select a weakly informative prior for $\eta$ that is skeptical of
large values given that $\nu$ is on the log-odds scale. For $\rho$ I
use an informative prior that constrains an overwhelming majority of
the prior probability mass to a *length-scale* between 4 and 10, which
provides a strong smoothing effect and aids in computation by avoiding
unreasonably large values far beyond the observed range
[@Betancourt2017].
\begin{align*}
    \eta & \sim \mathsf{HalfNormal}(0, 1) \\
    \rho & \sim \mathsf{InvGamma}(15.5031, 78.2992)
\end{align*}

@@

[^thresholds]: Thresholds relate to how the continuous latent trait
    for a given variable is mapped on to the ordinal answer categories
    presented in the survey. Due to differential item functioning,
    experts may perceive the magnitude of difference between answer
    items differently. For example, for the variable *v2elvotbuy*,
    election vote buying, experts may vary in how they interpret the
    difference between a 2 ("Restricted Vote Buying") and a 3 ("Almost
    No Vote Buying"). These two concepts, rater reliability and
    varying rater thresholds, correspond respectively to the
    *discrimination* and *difficulty* parameters in classical IRT
    models.

@@

In the case of the legislative variables there is the added issue of
missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves unbounded latent variables, minimal values cannot be
imputed without a nonlinear transformation to a bounded scale. I thus
take a cue from *v2x_horacc* and include into the model a dichotomized
version of *v2lgbicam*, how many chambers does the legislature
contain, denoted by the Bernoulli distributed variable, $w$, where
country-years with a sitting legislature are given a $1$ and
country-years where the legislature is missing or has been suspended,
$0$. The probability of a legislature being present, $\pi_i$, is then
a linear function of the latent factor, $\theta$, transformed by the
inverse logit such that $\pi_i \in [0, 1]$. The practical effect is
that where there is data for the legislative manifest variables, the
latent factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables. The resulting model specification, including
priors, can be summarised as follows:
\begin{align*}
    w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
    \pi_i & = \mathsf{logit}^{-1}(\eta + \delta \theta_i) \\
    \eta & \sim \mathsf{Normal}(0, 5) \\
    \delta & \sim \mathsf{LogNormal}(0, 1)
\end{align*}

@@

As mentioned previously, to control for time dependence --- past and
current civil conflicts can impact the present risk of conflict onset
as well as the current level of executive constraints -- I include the
number of peace years since the last conflict incident.[^time] I also
include a dummy variably for ongoing high intensity conflict since
devastating civil wars may have a particularly destabilizing
effect.[^intensity] To account for spillover effects, I include a
dummy variable for whether a neighbouring country is currently
experiencing a civil conflict.[^geo] Finally, one notable lack is a
general measure of political instability, which I omit since it could
be highly correlated with executive constraints and would potentially
bias results, at least for years of crisis where the risk of civil
conflict is highest. I do, however, include a dummy variable for the
independence year since newly sovereign states may be especially
vulnerable.

@@

Table \ref{tab:fa} and figure \ref{fig:fpars} summarise the posterior
draws for the measurement model parameters with 80 percent posterior
quantile intervals. According to the results, the factor loadings,
$\lambda$, for each manifest variable are fairly comparable, with the
exception of the lower value for *v2lgfunds* ("legislature controls
its own resources") and to a lesser extent *v2lgqstexp* ("legislature
questions executive branch officials in practice"). Relative to the
other manifest variables, both items are less informative in inferring
the values of the latent factor --- a $\lambda$ of zero would indicate
that the latent factor does not explain any of the variation in the
manifest variable. As for the variable specific intercepts, $\gamma$,
they are largely clustered around zero as to be expected given that
the manifest variables were standardized prior to the model run. Where
deviations do occur, most notably with several of the legislative
variables, it is because of the measurement error component of the
model shifting the mean for a specific variable away from zero. For
example, after adjusting for measurement error, the mean for
$v2lgfunds$ is on average slightly higher than the remaining manifest
variables, hence the negative intercept.

(ref:fa) Executive constraints measurement model parameter
    estimates with 80 percent posterior credible intervals.

@@

```{r fig.width = 6, fig.height = 4, fig.cap = "\\label{fig:theta} (ref:theta)"}
theta_summarised <- post_summarise(theta, probs = c(0.10, 0.5, 0.90)) %>%
    rename(codelow = `10%`, median = `50%`, codehigh = `90%`)

@@

(ref:tie) Executive constraints in 2000 versus 2010 for several
    selected countries.

```{r fig.cap = "\\label{fig:tie} (ref:tie)"}
targets <- c("Algeria", "Argentina", "Armenia", "Chile", "Colombia",
             "Egypt", "Georgia", "Ghana", "Hungary", "Indonesia",
             "Iran", "Israel", "Japan", "Kenya", "Mexico", "Poland",
             "Portugal", "Romania", "Rwanda", "Turkey", "Tunisia",
             "United States of America", "Zimbabwe")

@@

tfplot(df) + xlab("Theta")
```

@@

(ref:reg) Regression slopes with 80 percent posterior credible
    intervals and the posterior probability of each parameter being
    less than and greater than zero.

```{r reg}
reg <- read_post("posteriors/full_model/reg_posteriors.csv")
